---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Note:  When downloading the file with download.file, you have to change the https to http and use the mode="wb" option or the download.file will error out.

### Handling odd behavior with na.rm=TRUE
Normally you can use the dplyr summarise function to sum subsets of a dataframe, and use na.rm=TRUE to exclude NA values in the calculation.  There are certain dates (e.g. 2012-10-01) for which all intervals have steps=NA.  When computing a sum for the intervals on this day, with na.rm=TRUE, R should return NA.  Instead it returns 0.  This is incorrect.  To ensure that NA values do not skew calculations such as mean and median (which are run on the totalsteps/day), the rows with NA values must be removed using is.na function in the preprocessing stage. 

The dataframe no_na will contain the dataset where rows with steps=NA are removed.
The dataframe movement will contain the original dataset with NA values, as read from the .csv.

```{r loading and preprocessing}
download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip","activity.zip", mode="wb")
unzip("activity.zip")
movement<-read.csv("activity.csv", header=TRUE)
no_na<-movement[!is.na(movement$steps),]
library(dplyr)
library(ggplot2)
```


## What is mean total number of steps taken per day?

```{r total steps per day, fig.width=10}
bydate<-group_by(no_na, date)
total_bydate<-summarise(bydate, totalsteps=sum(steps))
g<-ggplot(total_bydate, aes(totalsteps))
g<-g+geom_histogram(fill="red", col="black", binwidth=5000)
print(g)
```

Below is the mean steps per day.
```{r report mean steps per day}
print(mean(total_bydate$totalsteps))
```
Below is the median steps per day.
```{r report median steps per day}
print(median(total_bydate$totalsteps))
```

## What is the average daily activity pattern?
The following computes the mean number of steps occurring in each 5 minute interval, looking across all days
```{r mean steps average per 5 minute interval}
by5min<-group_by(no_na, interval)
mean_by5min<-summarise(by5min, meansteps=mean(steps))
g<-ggplot(mean_by5min, aes(x=interval, y=meansteps)) + geom_line() + xlab("5 min interval") + ylab("mean steps/5 min interval")
print(g)
```

Note: we next find the 5 minute interval at which the maximum number of average steps occurs (looking across all days)
```{r find 5 minute interval at which maximum mean steps occurs}
mean_by5min[which.max(mean_by5min$meansteps),"interval"]
```

## Imputing missing values

The following computes the total number of missing values (NAs):
```{r compute number of NA values in steps}
na_val<-nrow(movement[is.na(movement$steps),])
print(na_val)
```
The total number of NA values in the steps column is `r na_val`

The following code demonstrates a strategy for replacing the NA values in the step column with a reasonable value for number of steps. Recall, we previously computed the mean number of steps for each 5 minute interval, across all days.  The code will substitute this value for any 5 minute interval showing NA. A new dataset called movement2 is created that contains the NA values replaced with the 5 minute interval mean across all days.

```{r strategy for substituting NA}
movement2<-movement
for (i in 1:nrow(movement2)){
    if (is.na(movement2$steps[i])){
        #replace this 5 minute interval with the average value of this same 5 min interval across all the days
        movement2$steps[i]=round(mean_by5min[mean_by5min$interval==movement2$interval[i], "meansteps"])
    }
}
```

The following computes the mean steps/day and median steps/day, using the new dataset movement2.  This dataset substitutes the avg 5 minute values (across all days) for intervals where there were NA values:
```{r total steps on dataset movement2, fig.width=10}
movement2$steps<-as.numeric(movement2$steps)
bydate_m2<-group_by(movement2, date)
total_bydate_m2<-summarise(bydate_m2, totalsteps_m2=round(sum(steps), 3))

g<-ggplot(total_bydate_m2, aes(totalsteps_m2))
g<-g+geom_histogram(fill="red", col="black", binwidth=5000)
print(g)
```

Below is the mean steps per day, using a substituted NA value.
```{r mean on dataset movement2}
print(mean(total_bydate_m2$totalsteps_m2))
```
Below is the median steps per day, using a substituted NA value.
```{r median on dataset movement2}
print(median(total_bydate_m2$totalsteps_m2))
```
The substituted values increased the number of days where total steps was in the range between 10,000 and 15,000 (from ~28 -> ~36).  The mean and median values are both lower than the corresponding values where the NA was removed.

## Are there differences in activity patterns between weekdays and weekends?
This is the code that adds a new column of type factor identifying a date as weekend or weekday.  Per the instruction, the dataset that is used is the one where NA values were filled in (the movement2 dataset).

```{r weekday or weekend}
movement3<-movement2 %>% mutate(typeofday=ifelse(grepl('Sat|Sun', weekdays(as.Date(date))), 'weekend', 'weekday'))
movement3$typeofday<-as.factor(movement3$typeofday)
```
The code below will group the rows by typeofday (weekend or weekday), then by interval.  Then a two panel plot will be made, with each panel showing the result for weekend vs weekday.
```{r plot weekend vs weekday}
by5min_typeofday<-group_by(movement3, typeofday, interval)
mean_by5min_typeofday<-summarise(by5min_typeofday, meansteps=round(mean(steps),3))
g<-ggplot(mean_by5min_typeofday, aes(x=interval, y=meansteps)) + geom_line() + xlab("5 min interval") + ylab("mean steps/5 min interval")
g<-g+facet_wrap(~typeofday,ncol=1)
print(g)
```

The charts do show a difference in activity levels between weekdays and weekend.  During weekdays there is a sharp spike in activity early in the morning.  This is probably due to this person excercising.  After that, the average activity level during a weekday is lower than that on a weekend.  This person may have a desk job which does not require much activity during the day.  During the weekend, the average level of activity during the day was higher than a weekday.  This is probably due to this person being out and about running errands and doing things outside.